---
title: "Analyzing State Data"
author: "Camryn Tidsworth"
date: "2024-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(tidyverse)
```

## Welcome

For this project, I decided to explore the publicly available dataset "state.x77." I explore the data, do some very cursory cleaning (as the data was already very clean to begin with), and find some interesting correlations to analyze.

First, let's learn about our data.

```{r}
?state.x77
```

According to the information in the help pane, this dataset is a "matrix with 50 rows and 8 columns giving the following statistics in the respective columns.

Population:

* population estimate as of July 1, 1975

Income:

* per capita income (1974)

Illiteracy:

* illiteracy (1970, percent of population)

Life Exp:

* life expectancy in years (1969–71)

Murder:

* murder and non-negligent manslaughter rate per 100,000 population (1976)

HS Grad:

* percent high-school graduates (1970)

Frost:

* mean number of days with minimum temperature below freezing (1931–1960) in capital or large city

Area:

* land area in square miles"

Now that we know some background about our data let's take a look at the first few columns.

```{r}
head(state.x77)
```

Next, we'll see if there are any missing values in our dataset that will need to be considered during the anlysis.

```{r}
sum(is.na(state.x77))
```
Good luck! There are no missing values in the data. 

Now let's get some initial summary statistics.

```{r}
summary(state.x77)
```

The mean life expectancy is lower than I would've expected. But then this data was collected in 1961-1971. Let's hope the life expectancy has increased since then.

I'm curious about the population density of each state, so I'll add a column with this new metric. In order to do this, I first need to convert state.x77 to a dataframe.

```{r}
state.stats <- state.x77 %>%
  as.data.frame(.)
```

Now we can make the new column.

```{r}
state.x77_pd <- mutate(state.stats, pop_density = Population / Area)
glimpse(state.x77_pd)
```

Now let's play with some scatter and line plots to see if there are any correlations in the data.

```{r}
ggplot(state.x77_pd, aes(x = Illiteracy,
                y = Income)) +
  geom_point() +
  geom_smooth(method = "lm") 
```

Illiteracy and income have a negative correlation.

```{r}
ggplot(state.x77_pd, aes(x = `Life Exp`,
                y = Income)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Life expectancy and income have a slight positive correlation. 

```{r}
ggplot(state.x77_pd, aes(x = Frost,
                y = Murder)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Interestingly, frost and murder are negatively correlated. This would be an interesting area for future analysis: does frost somehow cause murder, or is this a totally random and inconsequential correlation?

Finally, let's assume I have a client who is interested in the connection between litteracy rates and income. I'll create a summary table showing this relationship.

```{r}
state.x77_pd %>% 
  group_by(Illiteracy) %>% 
  summarise(mean(Income),
            median(Income))
```

Based on both the previous graph and this tibble, income and literacy have a positive linear correlation in the United States. 

## Conclusion

Thanks for taking a look at my project. As you can see, this was a cursory, semi_unfocused analysis project that I undertook to practice and demonstrate common R skills. In the future I look forward to doing more focused projects to answer specific questions. Maybe I'll dig into that frost-murder relationship...

As always, constructive feedback is very welcome!


